---
title: "Theory of PCA - Answer Key"
author:
  - Noor Sohail
  - Will Gammerdinger
date: "2025-09-05"
---

```{r}
#| label: load_data_R
#| echo: false
# Items to pre-load
library(tidyverse)

# Create a vector for Cell IDs
cells <- c("Cell_1", "Cell_2", "Cell_3", "Cell_4")
# Create a vector to hold expression values for Gene A across all of the cells
Gene_A <- c(0, 12, 65, 23)
# Create a vector to hold expression values for Gene B across all of the cells
Gene_B <- c(4, 30, 57, 18)

# Create a tibble to hold the cell names and expression values
expression_tibble <- tibble(cells, Gene_A, Gene_B)

# Determine the center of the data by:
# Finding the average expression of gene A
Gene_A_mean <- mean(expression_tibble$Gene_A)
# Finding the average expression of gene B
Gene_B_mean <- mean(expression_tibble$Gene_B)

# Create a vector to hold the center of the data
center_of_data <- c(Gene_A_mean, Gene_B_mean)
# Assign names to the components of the vector
names(center_of_data) <- c("Gene_A", "Gene_B")

# Shift the data points so that they data is centered on the origin
recentered_expression_tibble <- expression_tibble %>%
  mutate(
    Gene_A = Gene_A - Gene_A_mean,
    Gene_B = Gene_B - Gene_B_mean
  )

# Move the cell IDs to the rownames and convert the tibble to a matrix
recentered_expression_matrix <- recentered_expression_tibble %>% 
  column_to_rownames("cells") %>% 
  as.matrix()

# Create a covariance matrix
cov_matrix <- cov(recentered_expression_matrix)

# Find the eigenvalues and eigenvectors of the covariance matrix
eig <- eigen(cov_matrix)

# Transform the data into PC space by multiply the re-centered expression matrix by the eigenvectors
pc_scores <- recentered_expression_matrix %*% eig$vectors
# Name the columns in pc_scores object
colnames(pc_scores) <- c("PC_1", "PC_2")

# Calculate the percent of variance explained by each PC using the eigenvalues
pct_var_explained <- (eig$values / sum(eig$values)) * 100
# Name the elements of the pct_var_explained by their PC
names(pct_var_explained) <- c("PC_1", "PC_2")

# Create a tibble to hold the PC scores we found and also make the Cell IDs into a column
pc_scores_tibble <- pc_scores %>% 
  as.data.frame() %>% 
  rownames_to_column("cells") %>% 
  as_tibble()

# Run prcomp() on the expression tibble after moving the Cell IDs to be rownames
prcomp_PCA <- expression_tibble %>% 
  column_to_rownames("cells") %>% 
  prcomp()

# Print the eigenvalues found by prcomp() by squaring prcomp_PCA$sdev
prcomp_eigenvalues <- prcomp_PCA$sdev ** 2
# Name the elements of the prcomp_eigenvalues by their PC
names(prcomp_eigenvalues) <- c("PC_1", "PC_2")
```

```{python}
#| label: load_data_Python
#| echo: false
# Import libraries
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
from sklearn.decomposition import PCA

# Create a list for Cell IDs
cells = ["Cell_1", "Cell_2", "Cell_3", "Cell_4"]
# Create a list to hold expression values for Gene A across all of the cells
Gene_A = [0, 12, 65, 23]
# Create a list to hold expression values for Gene B across all of the cells
Gene_B = [4, 30, 57, 18]

# Create a DataFrame to hold the cell names and expression values
expression_df = pd.DataFrame({
    "cells": cells,
    "Gene_A": Gene_A,
    "Gene_B": Gene_B
}).set_index("cells") 

# Determine the center of the data by:
# Finding the average expression of gene A
Gene_A_mean = expression_df["Gene_A"].mean()
# Finding the average expression of gene B
Gene_B_mean = expression_df["Gene_B"].mean()

center_of_data = {
    "Gene_A": Gene_A_mean,
    "Gene_B": Gene_B_mean,
}

# Shift the data so it is centered on the origin
recentered_expression_df = expression_df.assign(
    Gene_A = expression_df["Gene_A"] - Gene_A_mean,
    Gene_B = expression_df["Gene_B"] - Gene_B_mean,
)

# Create a covariance matrix
cov_matrix = recentered_expression_df.cov()

# Find the eigenvalues and eigenvectors of the covariance matrix
eig_values, eig_vectors = np.linalg.eigh(cov_matrix)

# Sort the eigenvalues and eigenvectors by descending variance explained
# Indices for descending order
idx = np.argsort(eig_values)[::-1]
# Reordered eigenvalues
eig_values = eig_values[idx]
# Reordered eigenvectors
eig_vectors = eig_vectors[:, idx] 

# Place the eigenvectors into a dataframe
eig_vectors_df = pd.DataFrame(
    eig_vectors
)

pc_scores = recentered_expression_df.values.dot(eig_vectors)

pc_scores_df = pd.DataFrame(
    pc_scores,
    index = recentered_expression_df.index,
    columns = ["PC_1", "PC_2"]
)

# Calculate the percent of variance explained by each PC using the eigenvalues
pct_var_explained = (eig_values / eig_values.sum()) * 100

# Create PCA object
pca_sklearn = PCA()

# Run PCA on our expression data
pc_scores_sklearn = pca_sklearn.fit_transform(expression_df)

# Print the eigenvectors found by scikit-learn
eigenvector_sklearn = pd.DataFrame(
    pca_sklearn.components_.T
)

pc_scores_sklearn_df = pd.DataFrame(
    pc_scores_sklearn,
    index = expression_df.index,
    columns = ["PC_1", "PC_2"]
)
```

## Exercise 1

There are a couple properties of variance and co-variance that we can verify:

*cov(X,X)* is equal to *var(X)*. We can observe this is mathematically below:

$$
\operatorname{cov}(X, X)
= \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})(x_i - \bar{x})
= \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^{2}
= \operatorname{var}(X)
$$

As a result, you will sometimes see covariance matrices written as:

$$
\begin{bmatrix}
\operatorname{var}(X) & \operatorname{cov}(X, Y) & \dots & \operatorname{cov}(X, Z) \\
\operatorname{cov}(Y, X) & \operatorname{var}(Y) & \dots & \operatorname{cov}(Y, Z) \\
\dots & \dots & \dots & \dots \\
\operatorname{cov}(Z, X) & \operatorname{cov}(Z, Y) & \dots & \operatorname{var}(Z)
\end{bmatrix}
$$

1. Confirm this property by estimating the variance for Gene A.

::: {.panel-tabset group="language"}

### R

```{r}
#| label: covariance_check_1_R
# Estimate the variance for Gene A
var(recentered_expression_matrix[, "Gene_A"])
```

### Python

```{python}
#| label: covariance_check_1_Python
# Estimate the variance for Gene A
print(recentered_expression_df["Gene_A"].var())
```

:::

2. Now estimate the covariance for Gene A and Gene A

::: {.panel-tabset group="language"}

### R

```{r}
#| label: covariance_check_2_R
# Estimate the covariance for Gene A and Gene A
cov(recentered_expression_matrix[, "Gene_A"], recentered_expression_matrix[, "Gene_A"])
```

### Python

```{python}
#| label: covariance_check_2_Python
# Estimate the covariance for Gene A and Gene A
print(recentered_expression_df["Gene_A"].cov(recentered_expression_df["Gene_A"]))
```

:::

3. Is the value the same? Does it match the value in the covariance matrix for Gene A and Gene A?

::: {.panel-tabset group="language"}

### R

```{r}
#| label: covariance_matrix_check_R
# Extract the covariance estimate of Gene A and Gene A from the covariance matrix
cov_matrix["Gene_A", "Gene_A"]
```

### Python

```{python}
#| label: covariance_matrix_check_Python
# Extract the covariance estimate of Gene A and Gene A from the covariance matrix
print(cov_matrix.loc["Gene_A", "Gene_A"])
```

:::

4. *cov(X,Y)* is equal to *cov(Y,X)*. We can observe this is mathematically below:

$$
\operatorname{cov}(X, Y)
= \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})
= \frac{1}{n-1} \sum_{i=1}^{n} (y_i - \bar{y})(x_i - \bar{x})
= \operatorname{cov}(Y, X)
$$

Estimate the covariance of Gene B and Gene A.

::: {.panel-tabset group="language"}

### R

```{r}
#| label:  covariance_check_3_R
# Estimate the covariance of Gene B and Gene A
cov(recentered_expression_tibble[, "Gene_B"], recentered_expression_tibble[, "Gene_A"])
```

### Python

```{python}
#| label:  covariance_check_3_Python
# Estimate the covariance of Gene B and Gene A
print(recentered_expression_df["Gene_B"].cov(recentered_expression_df["Gene_A"]))
```

:::

5. How does this compare to the covariance that we estimated by hand?

It is the same.

## Exercise 2

When looking at the percent explained by each principal component, the first principal component should explain the most and each of the following principal components should explain less than the previous principal component. Let's have a look at our `pct_var_explained` object, are our results congruent with this expectation?

::: {.panel-tabset group="language"}

### R

```{r}
#| label: pct_explained_R
# Print the percent variance explained
pct_var_explained
```

### Python

```{python}
#| label: pct_explained_Python
# Print the percent variance explained
print(pct_var_explained)
```

:::

Yes, `PC_1` explains `python pct_var_explained[0]` and `PC_2` explains `python pct_var_explained[1]`

## Exercise 3

Create a plot of the Principal Components Analysis derived from `prcomp()`/`PCA()`. Is it the same as the plot we derived except only reflected over the x-axis (R)/rotated 180&deg;(Python)?

::: {.panel-tabset group="language"}

### R

```{r}
#| label: fig-plotting_PCA_prcomp_R
#| fig-cap: "Scatterplot of `PC_1` vs. `PC_2` from `prcomp()`."
# Create a tibble to hold the PC scores prcomp() found and also make the Cell IDs into a column
prcomp_pc_scores_tibble <- prcomp_PCA$x %>% 
  as.data.frame() %>% 
  rownames_to_column("cells") %>% 
  as_tibble()

# Plot the PC scores found by prcomp()
ggplot(prcomp_pc_scores_tibble, aes(x = PC1, y = PC2, label = cells)) +
  geom_point( color = "cornflowerblue") +
  geom_text(hjust = 0, vjust = -1) +
  theme_bw() +
  xlim(-50, 50) +
  ylim(-12, 12) +
  xlab(paste0("PC 1 (Variance Explained ", round(prcomp_eigenvalues["PC_1"]/sum(prcomp_eigenvalues) * 100, digits = 2),"%)")) +
  ylab(paste0("PC 2 (Variance Explained ", round(prcomp_eigenvalues["PC_2"]/sum(prcomp_eigenvalues) * 100, digits = 2),"%)")) +
  ggtitle("PCA of Expression Values from Four Cells") +
  theme(plot.title = element_text(hjust = 0.5))
```

Yes, it is the same plot just reflected over the x-axis.

### Python

```{python}
#| label: plotting_PCA_PCA_Python_show
#| eval: false
#| echo: true
sns.set(style="whitegrid")

# Initialize a plot with a specific size
plt.figure(figsize = (8, 6))

# Add a scatterplot layer to the plot, coloring points by genotype
PCA_sklearn_plot = sns.scatterplot(data = pc_scores_sklearn_df,
                x = "PC_1", 
                y = "PC_2")

# Set x-axis limits
plt.xlim(left = -55,
         right = 55)

# Set y-axis limits
plt.ylim(bottom = -15,
         top = 15)

# Add cell labels at PC coordinates
for cell_id, row in pc_scores_sklearn_df.iterrows():
    PCA_sklearn_plot.text(
        x = row["PC_1"] + 1,
        y = row["PC_2"] + 1,
        s = cell_id,
        fontsize = 9,
        ha = "left",
        va = "bottom"
    )

# Change the text of the x-axis label
plt.xlabel(xlabel = f"PC 1 (Variance Explained {pca_sklearn.explained_variance_ratio_[0].round(4) * 100}%)")

# Change the text of the y-axis label
plt.ylabel(ylabel = f"PC 2 (Variance Explained {pca_sklearn.explained_variance_ratio_[1].round(4) * 100}%)")

# Add plot title
plt.title(label = "PCA of Expression Values from Four Cells")

# Render the plot
plt.show()
```

```{python}
#| label: fig-plotting_PCA_PCA_Python_hidden
#| fig-cap: "Scatterplot of `PC_1` vs. `PC_2` from scikit-learn."
#| echo: false

def plot_pca_sklearn(pc_scores_sklearn_df, pca_sklearn):
    sns.set(style="whitegrid")

    # Initialize a plot with a specific size
    fig, ax = plt.subplots(figsize=(8, 6))

    # Add a scatterplot layer to the plot
    sns.scatterplot(
        data=pc_scores_sklearn_df,
        x="PC_1",
        y="PC_2",
        ax=ax
    )

    # Set axis limits via the Axes object
    ax.set_xlim(-55, 55)
    ax.set_ylim(-15, 15)

    # Add cell labels at PC coordinates
    for cell_id, row in pc_scores_sklearn_df.iterrows():
        ax.text(
            x=row["PC_1"] + 1,
            y=row["PC_2"] + 1,
            s=cell_id,
            fontsize=9,
            ha="left",
            va="bottom"
        )

    # Axis labels with variance explained from sklearn PCA
    ax.set_xlabel(
        f"PC 1 (Variance Explained {pca_sklearn.explained_variance_ratio_[0].round(4) * 100}%)"
    )
    ax.set_ylabel(
        f"PC 2 (Variance Explained {pca_sklearn.explained_variance_ratio_[1].round(4) * 100}%)"
    )

    # Title
    ax.set_title("PCA of Expression Values from Four Cells")

    return fig, ax

fig, ax = plot_pca_sklearn(pc_scores_sklearn_df, pca_sklearn)
plt.show()
```

Yes, it is the same plot just rotated 180&deg;.

:::