---
title: "Clustering Quality Control - Answer Key"
author:
  - Noor Sohail
  - Will Gammerdinger
date: "2026-08-18"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| label: load_libraries_data_R
#| echo: false
# Load libraries and data
library(Seurat)
library(tidyverse)
library(ggplot2)
library(cowplot)

# Load data
seurat_integrated <- readRDS("data/additional_data/seurat_integrated.rds")
Idents(seurat_integrated) <- "integrated_snn_res.0.8"

# Set randomness
set.seed(1234)
```

```{python}
#| label: load_libraries_data_py
#| echo: false
# Load libraries and data
import scanpy as sc
import numpy as np
import seaborn as sns
import matplotlib.pyplot as plt

# Set tight_layout universally
plt.rcParams["figure.autolayout"] = True
plt.rcParams["figure.constrained_layout.use"] = True

# Load data
adata_integrated = sc.read_h5ad("data/additional_data/adata_integrated.h5ad")

# Set randomness
np.random.seed(1234)
```

# Exercise 1



1. Hypothesize the clusters corresponding to each of the different clusters in the table:

| Cell Type | Clusters |
|:---:|:---:|
| CD14+ monocytes | 2, 4 | 
| FCGR3A+ monocytes | 10 |
| Conventional dendritic cells | 13 |
| Plasmacytoid dendritic cells | 16 |
| Marcrophages | - |
| B cells | 7, 11, 14 |
| T cells | 1, 3, 6 |
| CD4+ T cells | 1, 3, 6 |
| CD8+ T cells| 5 |
| NK cells | 9, 12 |
| Megakaryocytes | 15 |
| Erythrocytes | - |


::: {.panel-tabset group="language"}
## R

```{r}
#| label: fig-DotPlot_known_markers
#| fig-cap: DotPlot representing top marker genes for a variety of cell types, with each circle representing the average expression of that cluster and the size showing the percentage of cells that express that gene.
#| fig.width: 20
# List of known celltype markers
markers <- list(
  "CD14+ monocytes" = c("CD14", "LYZ"),
  "FCGR3A+ monocytes" = c("FCGR3A", "MS4A7"),
  "Conventional dendritic cells" = c("FCER1A", "CST3"),
  "Plasmacytoid dendritic cells" = c("IL3RA", "GZMB", "SERPINF1", "ITM2C"),
  "B cells" = c("CD79A", "MS4A1"),
  "T cells" = c("CD3D"),
  "CD4+ T cells" = c("IL7R", "CCR7"),
  "CD8+ T cells" = c("CD8A"),
  "NK cells" = c("GNLY", "NKG7"),
  "Megakaryocytes" = c("PPBP"),
  "Erythrocytes" = c("HBB", "HBA2")
)

# Create dotplot based on RNA expression
DotPlot(seurat_integrated, markers, assay = "RNA") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r}
#| label: celltype_dimPlot
# Iterate through each celltype
for (celltype in names(markers)) {
  # Grab genes for each celltype
  genes <- markers[[celltype]]
  
  # UMAP plot
  p <- FeaturePlot(seurat_integrated,
                   reduction = "umap",
                   features = genes,
                   order = TRUE,
                   min.cutoff = "q10",
                   label = TRUE) + 
        ggtitle(celltype)
  
  print(p)
}
```

## Python

```{python}
#| label: fig-DotPlot_known_markers_py
#| fig-cap: DotPlot representing top marker genes for a variety of cell types, with each circle representing the average expression of that cluster and the size showing the percentage of cells that express that gene.
#| fig.width: 15
# List of known celltype markers
markers = {
    "CD14+ monocytes": ["CD14", "LYZ"],
    "FCGR3A+ monocytes": ["FCGR3A", "MS4A7"],
    "Conventional dendritic cells": ["FCER1A", "CST3"],
    "Plasmacytoid dendritic cells": ["IL3RA", "GZMB", "SERPINF1", "ITM2C"],
    "B cells": ["CD79A", "MS4A1"],
    "T cells": ["CD3D"],
    "CD4+ T cells": ["CD3D", "IL7R", "CCR7"],
    "CD8+ T cells": ["CD3D", "CD8A"],
    "NK cells": ["GNLY", "NKG7"],
    "Megakaryocytes": ["PPBP"],
    "Erythrocytes": ["HBB", "HBA2"],
}
sc.pl.dotplot(adata_integrated, 
              markers, 
              groupby = "leiden_0.8")
```


```{python}
#| label: celltype_embedding
for celltype, genes in markers.items():
    # Include cluster column + markers
    colors = ["leiden_0.8"] + genes

    # UMAP plot
    sc.pl.embedding(adata_integrated,
                    basis = "umap_scvi",
                    color = colors,
                    legend_loc = "on data",
                    ncols = 2,
                    show = False)

    plt.suptitle(celltype)
    plt.tight_layout()
    plt.show()
```

Based on the figures from above, we can roughly say that the cluster map to these celltypes:

| Cell Type | Clusters |
|:---:|:---:|
| CD14+ monocytes | 2 | 
| FCGR3A+ monocytes | 13 |
| Conventional dendritic cells | 3 |
| Plasmacytoid dendritic cells | 5 |
| Macrophages | - |
| B cells | 10, 11 |
| T cells | 0, 1, 6, 8 |
| CD4+ T cells | 0, 1, 4, 6 |
| CD8+ T cells| 8 |
| NK cells | 9 |
| Megakaryocytes | 15 |
| Erythrocytes | - |

:::

***

[Back to Lesson >>](11_clustering_quality_control.qmd)

[Back to Schedule](../schedule/schedule.qmd)